---
title: "Column Mapping, Config & Data Format"
description: >
  How to adapt splsleep to your own dataset: YAML column mapping, threshold
  configuration, timestamp format, input data structure, and manual
  correction CSV templates.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Column Mapping, Config & Data Format}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

splsleep is fully configurable via a YAML configuration file: map your
dataset's column names to pipeline internals and adjust thresholds without
modifying any R code.

```{r setup, eval = FALSE}
library(splsleep)
```

## Install and run

```r
# Install from GitHub
renv::install("cyracaid/sleepdiary-cleaner")

# Load and run
library(splsleep)
run_pipeline()
```

## Using with your own dataset

The pipeline is fully configurable via a YAML configuration file. This lets
you map your dataset's column names to pipeline variables and adjust
thresholds without modifying any R code.

```r
# Step 1: Copy the configuration template
library(splsleep)
file.copy(system.file("config_template.yaml", package = "splsleep"),
          "my_study.yaml")
```

**Step 2: Edit `my_study.yaml`**

The file starts with the only two things you must change:

```yaml
data:
  files:
    main: "your_data.rds"        # Your sleep diary file (.rds or .csv)
    extra: ""                    # Leave empty unless StartDate lives in a separate file
```

Three common scenarios:
- **Everything in one file** (most datasets): `main: "my_data.rds"`, `extra: ""`
- **Data split across two files**: `main: "ema_vars.rds"`, `extra: "dates.csv"`
- **Your data is a CSV**: `main: "my_data.csv"`, `extra: ""` — the `.csv` extension is auto-detected

### Column Mapping

Map your dataset's column names to the pipeline's internal variables:

```yaml
column_mapping:
  identifiers:
    pid: "subject_id"          # your participant ID column
    day_num: "study_day"       # your day number column
  timestamp:
    time_bed_hhmm: "bedtime"   # your bedtime HH:MM column
    time_bed_ampm: "bed_ampm"  # your bedtime AM/PM column
    time_sleep_hhmm: "sleeptime"
    time_sleep_ampm: "sleep_ampm"
  duration:
    sol: "sleep_onset_latency" # your SOL column (minutes)
    waso: "wake_after_onset"   # your WASO column (minutes)
  substance:
    caffeine: "caffeine_cups"
    alcohol: "alcohol_drinks"
```

### Thresholds

Adjust detection sensitivity for your study population:

```yaml
classification:
  metric_validation:
    sol:
      excessive_minutes: 120   # SOL > 2h → flagged
    se:
      min_valid_percent: 0
      max_valid_percent: 100
    tst_tib_ratio:
      min_ratio: 0.5
      max_ratio: 1.0
  flag_severity:
    poor_efficiency_threshold_pct: 70   # SE < 70% → flag
    high_sol_threshold_hours: 1         # SOL > 1h → flag
    high_waso_threshold_hours: 1.5      # WASO > 1.5h → flag
```

Threshold rationale lives in `THRESHOLDS.md` — defaults are deliberately
lenient for healthy-adult samples; revisit for clinical populations.

### Timestamp Format

```yaml
timestamp:
  input_format: "hh:mm AM/PM"   # or "HH:MM", "HH:MM:SS"
  ampm:
    enabled: true
    pm_keywords: ["PM", "pm"]
```

**Step 3: Run with your configuration**

```r
run_pipeline(config = "my_study_config.yaml")
```

All pipeline scripts automatically read the config; no R code changes needed.

## Input data structure

**This repository contains no raw participant data.** All CSV files with
participant data are gitignored. Templates with synthetic data live in
`templates/`.

| Column group | Variables | Description |
|---|---|---|
| Identifiers | pid, day_num, row_id, participant | Participant and record IDs |
| Date | StartDate | Calendar date of the EMA session |
| Raw timestamps (HH:MM) | time_bed_am_hhmm, time_sleep_am_hhmm, time_awake_am_hhmm, time_getup_am_hhmm | Self-reported bed/sleep/awake/getup clock times |
| Raw timestamps (AM/PM) | time_bed_am_ampm, time_sleep_am_ampm, time_awake_am_ampm, time_getup_am_ampm | AM/PM indicator for each timestamp |
| Raw durations | duration_totalmin_sol_estimate_am, duration_totalmin_waso_estimate_am | Self-reported SOL and WASO in minutes |
| Nap/Exercise | duration_totalmin_napstoday_PM, exercise_PM_totalmin_[Light\|Moderate\|Vigorous\|Strength] | Self-reported nap and exercise durations |
| Substance use | caffeinetoday_PM_NumCaffeinatedDrinksSnacks_1, alcoholtoday_PM_NumAlcoholicDrinks_1, nicotine_amount_pm_doses, cannabis_amount_pm_doses | Self-reported substance use |
| WASO count | num_waso_estimate_am, num_waso_am | Number of wake bouts |

## Manual correction CSV templates

| Template File | Live File | Purpose |
|---|---|---|
| `templates/template_manual_error_corrections.csv` | `manual_error_corrections.csv` | Timestamp corrections (AM/PM, order) |
| `templates/template_manual_unusual_corrections.csv` | `manual_unusual_corrections.csv` | Accepted unusual patterns |
| `templates/template_manual_nap_exercise_corrections.csv` | `manual_nap_exercise_corrections.csv` | Nap/exercise duration corrections |
| `templates/template_manual_sleep_metric_duration_corrections.csv` | `manual_sleep_metric_duration_corrections.csv` | SOL/WASO metric corrections |
| `templates/template_manual_metric_review_acceptances.csv` | `manual_metric_review_acceptances.csv` | Human-accepted metric flags |
| `templates/template_second_review_checklist.csv` | `second_review_checklist.csv` | Second-person verification decisions |

## Output

| File | Contents |
|------|----------|
| `output/correction_status_final.csv` | Per-run summary: n_total, tst, sol, error/corrected/flag counts |
| `output/appendix_step_ledger.csv` | Per-step flag tracking ledger |
| `output/flagged_records_self_reported.csv` | Records flagged as SELF_REPORTED_FLAG |
| `latest_visualization_*/figure_index.png` | Contact-sheet index of all generated figures |
| `output/verification/real_n13990/`, `verification/synth_n280/` | Stable, never-overwritten verification artifacts |

Key rules the output structure encodes:
- `latest_visualization_<tag>_n<rows>/` is "latest", not "history" — wiped on
  every visualization run.
- `verification/<tag>_n<rows>/` is a sibling that is never touched by the wipe.
- Real-tagged output only ever lands under `output/` (gitignored); synthetic
  output routes outside `output/`.

## 中文 — 列映射、配置与数据格式

splsleep 通过 YAML 配置完全可配置：把数据集的列名映射到管线内部变量、调整
阈值，无需改 R 代码。

```r
# 复制配置模板
file.copy(system.file("config_template.yaml", package = "splsleep"), "my_study.yaml")
```

必须改的两项：`data.files.main`（你的数据文件 .rds/.csv）和 `data.files.extra`
（StartDate 单独文件时填，否则留空）。三种常见场景：单文件、双文件、CSV
（自动检测扩展名）。

**列映射**：`column_mapping` 段把标识符（pid/day_num）、时间戳（bed/sleep 的
HH:MM + AM/PM 列）、时长（SOL/WASO 分钟列）、物质使用列映射到管线内部名。

**阈值**：`classification` 段调整检测灵敏度。默认值故意宽松（健康成人样本）；
临床人群需重审（依据见 THRESHOLDS.md）。

**时间戳格式**：`timestamp.input_format` 支持 "hh:mm AM/PM" / "HH:MM" /
"HH:MM:SS"。

**运行**：`run_pipeline(config = "my_study.yaml")` — 所有脚本自动读配置。

**输入数据结构**：标识符、日期、原始时间戳（HH:MM + AM/PM 双列）、原始时长
（SOL/WASO 分钟）、小睡/运动、物质使用、WASO 次数。仓库不含真实数据，模板
用合成数据（`templates/`）。

**人工修正 CSV**：6 种模板（时间戳修正、异常接受、小睡/运动、SOL/WASO 指标
修正、指标接受、二次验证清单），模板用合成数据。

**输出**：`correction_status_final.csv`（运行摘要）、`appendix_step_ledger.csv`
（逐步标记账本）、`flagged_records_self_reported.csv`、图表索引、稳定验证产物
（snapshot/Bland-Altman/阈值验证，永不被覆盖）。`latest_visualization_*` 每次
覆盖；`verification_*` 永不触碰。
