---
title: "Interpreting the Pipeline Output"
description: >
  How to read correction_status_final.csv, the step flag ledger, regression
  checks, and the diagnostic figures — what each output means and what to
  check for.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Interpreting the Pipeline Output}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

After `run_pipeline()` finishes, two CSV files tell you everything. This
vignette explains how to read them, how to regression-check against a
previous run, and how to read the figures.

```{r setup, eval = FALSE}
library(sleepcleanr)
```

## Terminology

Sleep-diary metric abbreviations used throughout this vignette: **SOL**
(Sleep Onset Latency), **TST** (Total Sleep Time), **WASO** (Wake After
Sleep Onset), **SE** (Sleep Efficiency).

## 1. `output/correction_status_final.csv` — The Run Summary (Open This First)

One row per pipeline run. It answers: *"did the cleaning work as expected?"*

```r
read.csv("output/correction_status_final.csv")
```

| Column | It tells you... | Check this |
|--------|----------------|------------|
| `n_total` | Total records in your data | Must equal your input row count. If smaller, records were dropped somewhere. |
| `tst_mean_h` | Mean total sleep time in hours | 6.0–8.5 h is normal for most adult studies. If < 5 or > 10, something is off with the timestamp parsing or the study population is unusual. |
| `sol_mean_min` | Mean sleep onset latency in minutes | 10–45 min is normal. If > 60, either the population has high insomnia or AM/PM confusion was not fully corrected. |
| `n_clean` | Records that passed every check | Should be stable across runs (same data = same count). |
| `n_error` | Records with impossible temporal order (e.g., getup before bedtime) | Should be < 1% of total. If > 5%, review the survey design or data collection. |
| `n_corrected` | Records manually corrected via your CSV files | Should match the number of rows in your `manual_error_corrections.csv`. |
| `timestamp_issue` | Timestamps that could not be parsed into a valid time | 0 is normal. > 0 means some participants entered non-standard time formats. |
| `duration_issue` | Sleep metrics (SOL, SE, TST) outside configured thresholds | Small numbers are normal. If very large, your thresholds may be too strict or the data has quality problems. |
| `amount_flag` | Substance-use entries with unusual values | Should be 0 or very low. |
| `self_reported_flag` | Records where self-reported SOL/WASO disagrees with the computed value | Indicates perception bias. Check Figure 20 (SOL Perception Bias). |

**Stability rule:** Run twice on the same data → every number must be
identical. If not, something is non-deterministic.

## 2. `output/step_flag_ledger.csv` — The Per-Step Flag Tracker (Open Second)

One row per step × per standard × per category. It answers: *"at which step
did which flag appear, and did it persist?"*

```r
ledger <- read.csv("output/step_flag_ledger.csv")
library(dplyr)
ledger %>% filter(!is.na(count)) %>% arrange(step_id, standard)
```

Each row answers: *"at this step, using this standard, how many records fell
into this category?"*

### Column layout

| Column | What it is |
|--------|-----------|
| `step_id` | Pipeline step number |
| `label` | Human-readable step name |
| `n_total` | Total records in the pipeline at this point |
| `standard` | The evaluation system being tracked (see below) |
| `category` | The specific category within that standard |
| `count` | Number of records in this category (NA = not yet computed at this step) |
| `n_corrected` | Records manually corrected at this step |

The ledger uses 5 independent evaluation systems:

| Standard | First step with numbers | What it evaluates | Key categories |
|----------|:----------------------:|-------------------|----------------|
| `field_misentry` | 1.5 | Whether a duration estimate (SOL, WASO) exactly matches a timestamp — may indicate "typed in wrong box" | `none`, `SOL=time_sleep`, `SOL=time_bed`, `WASO=time_awake`, `WASO=time_getup` |
| `data_category` | 4 | Temporal order and plausibility of the bed → sleep → awake → getup sequence | `clean`, `error`, `unusual`, `equal_time_ok`, `skipped_na` |
| `flag_severity` | 7 | How many derived-metric flags (low SE, high SOL, high WASO) each record triggered | `Clean`, `Minor (1 flag)`, `Major (2+ flags)` |
| `duration_extreme` | 7 | Total sleep time outside physiologically plausible bounds | `OK`, `Too short (< 3 h)`, `Too long (> 12 h)` |
| `checkforerrors` | 8 | Summary of all auto-detection flags assembled in Step 8 | `TIMESTAMP_ISSUE`, `DURATION_ISSUE`, `AMOUNT_FLAG`, `SELF_REPORTED_FLAG` |

**Conceptual rule:** Standards are computed once and never re-computed. If a
standard's counts change after its first computation step, something is wrong.

**Validation rules:**

1. `field_misentry` — populated from Step 1.5 onward. Any `SOL=time_bed` or
   `WASO=time_getup` with `count > 0` = potential cross-field contamination.
2. `data_category` — from Step 6 onward the numbers must be **stable**:
   `equal_time_ok + skipped_na = n_total`.
3. `flag_severity` — from Step 7 onward identical across Steps 7, 8, 8.5:
   `Clean + Minor + Major = n_total - skipped_na`.
4. `duration_extreme` — `Too short + Too long` should be < 5% of `n_total`.
5. `checkforerrors` — populated at Step 8 only.

**Example (synthetic data, 280 rows):**

```
Step 7 (Compute metrics):
  data_category:    equal_time_ok = 266, skipped_na = 14        266 + 14 = 280 ✓
  flag_severity:    Clean = 251, Minor = 28, Major = 1         251 + 28 + 1 = 280 - 14 ✓
  duration_extreme: OK = 262, Too short = 1, Too long = 0
```

## 3. Regression Check (compare against a previous run)

`output/correction_status_old.csv` is not written by any pipeline script —
you create it yourself as a saved baseline before rerunning:

```r
# BEFORE rerunning: save the current output as your baseline.
file.copy("output/correction_status_final.csv", "output/correction_status_old.csv",
          overwrite = TRUE)
# ... rerun the pipeline here ...
# AFTER rerunning: compare against the baseline.
old <- read.csv("output/correction_status_old.csv")
new <- read.csv("output/correction_status_final.csv")
identical(old$tst_mean_h, new$tst_mean_h)
identical(old$sol_mean_min, new$sol_mean_min)
identical(old$n_clean, new$n_clean)
```

If they differ and the input data did not change, the pipeline output has
changed. Investigate.

## 4. Quick Reference Card

| Check | What to run | Pass if |
|-------|-------------|---------|
| Pipeline finished | `file.exists("output/correction_status_final.csv")` | `TRUE` |
| Reasonable TST | `tst_mean_h` between 6–8.5 | Yes |
| Reasonable SOL | `sol_mean_min` between 10–45 | Yes |
| Few errors | `n_error < 0.01 * n_total` | Yes |
| data_category stable | Counts identical across Steps 6–8.5 | Yes |
| flag_severity stable | Counts identical across Steps 7–8.5 | Yes |
| All records accounted | `equal_time_ok + skipped_na = n_total` | Yes |
| Deterministic | Same input → same output every time | Yes |

## 5. How to Read the Figures

Figures are saved to `latest_visualization_<tag>_n<rows>/` (overwritten each
run — no history). A `figure_index.png` contact sheet shows all figures at a
glance. Stable verification artifacts (snapshots, Bland-Altman plots,
threshold validation) live separately in `output/verification/<tag>/`.

**Publication figures (for a Methods section):**

| Figure | File | What it shows |
|--------|------|---------------|
| **Figure 1 — Pipeline Flow** | `pipeline_cleaning/01_Pipeline_Flow_Diagram.png` | Vertical flow diagram: raw → parsed → algo-corrected → manual-corrected → final valid, with counts and % at each stage |
| **Figure 2 — Correction Impact** | `research_ready/02_Correction_Impact.png` | A/B delta lollipops (only modified records, TST & SOL) + identity scatter + before/after summary table |

**Five must-check figures:**

**Five must-check figures:**

| Step | Figure (card) | Should look like | If not? |
|:----:|--------|------------------|---------|
| 1 | **01 Pipeline Record Flow** (§6.2) | Flow narrows gently; Clean dominates; Error + Unusual small (< 5%) | Spike at 0 or huge Error/Unusual share → parsing/AM-PM failure upstream |
| 2 | **A1 Step Flag Ledger** (§6.2) | Corrected bar appears ONLY at Step 6.5, flat after | Change after Step 6.5 = instability
| 3 | **18 Auto-Detected Dashboard** (§6.2) | Flag counts match `correction_status_final.csv` | Mismatch = misalignment
| 4 | **02B Distribution of Sleep Variables** (§6.2) | TST peaks 6–8 h, SOL right-skewed, WASO < 60, SE > 85% | SOL flat/bimodal = AM/PM confusion
| 5 | **13 Error Category Distribution** (§6.2) | Most records Clean/Minor; Error+Unusual < 5% | High = review manual CSVs

## 6. Figure-by-Figure Reference

Every figure the pipeline produces is documented as a self-contained card below: what each part means, the exact rules it is built from, what healthy data look like, what anomalies mean, and a paper-ready caption. Figures land in `latest_visualization_<tag>_n<rows>/pipeline_cleaning/` (diagnostic) and `.../research_ready/` (publication).

### 6.1 Figure Index

| # | File | Paper use | Code source | Card | Last checked |
|---|------|-----------|-------------|------|--------------|
| **01** | `pipeline_cleaning/01_Pipeline_Flow_Diagram.png` | Methods Fig 1 | `sleep_visualization.R:832` | [§6.2 · Pipeline Record Flow](#01-pipeline-record-flow) | 2026-09-25 |
| **02** | `research_ready/02_Correction_Impact.png` | Methods Fig 2 | `sleep_visualization.R:974` | [§6.2 · Correction Impact](#02-correction-impact) | 2026-09-25 |
| **02B** | `research_ready/02B_Distribution_Sleep_Variables.png` | Results | `sleep_visualization.R:1018` | [§6.2 · Distribution of Sleep Variables](#02b-distribution-of-sleep-variables) | 2026-09-25 |
| **03** | `research_ready/03_Sleep_Duration_Distribution.png` | Results | `sleep_visualization.R:1056` | [§6.2 · TST Distribution](#03-tst-distribution) | 2026-09-25 |
| **04** | `research_ready/04_Sleep_Duration_vs_Time_in_Bed.png` | Results | `sleep_visualization.R:1111` | [§6.2 · Sleep Duration vs Time in Bed](#04-sleep-duration-vs-time-in-bed) | 2026-09-25 |
| **04B** | `research_ready/04B_SOL_vs_Sleep_Duration.png` | Results | `sleep_visualization.R:1161` | [§6.2 · SOL vs Sleep Duration](#04b-sol-vs-sleep-duration) | 2026-09-25 |
| **05** | `research_ready/05_Variability_Sleep_Variables.png` | Results | `sleep_visualization.R:1211` | [§6.2 · Variability of Sleep Variables](#05-variability-of-sleep-variables) | 2026-09-25 |
| **06** | `pipeline_cleaning/06_Sleep_Duration_Post_Correction.png` | Supplement | `sleep_visualization.R:1285` | [§6.2 · Sleep Duration Post-Correction](#06-sleep-duration-post-correction) | 2026-09-25 |
| **07** | `pipeline_cleaning/07_Flag_Composition_Stacked.png` | Supplement | `sleep_visualization.R:1350` | [§6.2 · Flag Composition Stacked](#07-flag-composition-stacked) | 2026-09-25 |
| **09** | `research_ready/09_Bedtime_vs_Getup_Distribution.png` | Results | `sleep_visualization.R:1404` | [§6.2 · Bedtime vs Get-up Distribution](#09-bedtime-vs-get-up-distribution) | 2026-09-25 |
| **10** | `pipeline_cleaning/10_Extreme_Sleep_Duration.png` | Supplement | `sleep_visualization.R:1460` | [§6.2 · Extreme Sleep Durations](#10-extreme-sleep-durations) | 2026-09-25 |
| **13** | `pipeline_cleaning/13_Error_Category_Distribution.png` | Supplement | `sleep_visualization.R:1734` | [§6.2 · Error Category Distribution](#13-error-category-distribution) | 2026-09-25 |
| **13B** | `pipeline_cleaning/13B_Adjacent_Timestamp_Gaps.png` | Supplement | `sleep_visualization.R:1817` | [§6.2 · Adjacent Timestamp Gaps](#13b-adjacent-timestamp-gaps) | 2026-09-25 |
| **13C** | `pipeline_cleaning/13C_Detection_Outcomes_Heatmap.png` | Supplement | `sleep_visualization.R:1876` | [§6.2 · Detection Outcomes Heatmap](#13c-detection-outcomes-heatmap) | 2026-09-25 |
| **13D** | `pipeline_cleaning/13D_Threshold_vs_Noise_Ratio.png` | Supplement | `sleep_visualization.R:1920` | [§6.2 · Threshold vs Measurement Noise](#13d-threshold-vs-measurement-noise) | 2026-09-25 |
| **17** | `pipeline_cleaning/17_Top_Participants_Flags.png` | Supplement | `sleep_visualization.R:2222` | [§6.2 · Top Participants by Flag Rate](#17-top-participants-by-flag-rate) | 2026-09-25 |
| **18** | `pipeline_cleaning/18_Auto_Detected_Dashboard.png` | Supplement | `sleep_visualization.R:2322` | [§6.2 · Auto-Detected Dashboard](#18-auto-detected-dashboard) | 2026-09-25 |
| **20** | `research_ready/20_SOL_Perception_Bias.png` | Results | `sleep_visualization.R:2547` | [§6.2 · SOL Perception Bias](#20-sol-perception-bias) | 2026-09-25 |
| **20B** | `research_ready/20B_WASO_Perception_Bias.png` | Results | `sleep_visualization.R:2602` | [§6.2 · WASO Perception Bias](#20b-waso-perception-bias) | 2026-09-25 |
| **21** | `research_ready/21_Substance_Use_Availability.png` | Supplement | `sleep_visualization.R:2674` | [§6.2 · Substance Use Availability](#21-substance-use-availability) | 2026-09-25 |
| **22** | `research_ready/22_Substance_Use_Distribution.png` | Supplement | `sleep_visualization.R:2777` | [§6.2 · Substance Use Value Distribution](#22-substance-use-value-distribution) | 2026-09-25 |
| **23** | `research_ready/23_Caffeine_Consumption.png` | Supplement | `sleep_visualization.R:2839` | [§6.2 · Caffeine Consumption](#23-caffeine-consumption) | 2026-09-25 |
| **24** | `research_ready/24_Alcohol_Consumption.png` | Supplement | `sleep_visualization.R:2874` | [§6.2 · Alcohol Consumption](#24-alcohol-consumption) | 2026-09-25 |
| **A1** | `pipeline_cleaning/A1_Step_Flag_Ledger.png` | Supplement | `figure12_step_flag_table.R:15` | [§6.2 · Step Flag Ledger](#a1-step-flag-ledger) | 2026-09-25 |
| **P26** | `pipeline_cleaning/P26_PerParticipant_Flag_Rate.png` | Supplement | `sleep_visualization.R:2492` | [§6.2 · Participants Worth a Second Look](#p26-participants-worth-a-second-look) | 2026-09-25 |
| **R25** | `research_ready/R25_Sleep_Regularity_Weekday_Weekend.png` | Results | `sleep_visualization.R:2937` | [§6.2 · Sleep Regularity — Weekday vs Weekend](#r25-sleep-regularity--weekday-vs-weekend) | 2026-09-25 |
| **R26** | `research_ready/R26_Sleep_Composition_TIB_Breakdown.png` | Results | `sleep_visualization.R:2987` | [§6.2 · Sleep Composition — TIB Breakdown](#r26-sleep-composition--tib-breakdown) | 2026-09-25 |
| **R27** | `research_ready/R27_Sleep_Metrics_Correlation_Matrix.png` | Results | `sleep_visualization.R:3016` | [§6.2 · Sleep Metrics Correlation Matrix](#r27-sleep-metrics-correlation-matrix) | 2026-09-25 |

The retired figures `08_Sleep_Duration_by_Category` and `12_Pipeline_Correction_Progress` are no longer produced (their questions are answered by Figure 07 and A1); `11`, `14`, `15` and `16` are intentionally skipped on real data (empty/blank by design).

### 6.2 Figure Cards

#### 01 · Pipeline Record Flow

**File** `pipeline_cleaning/01_Pipeline_Flow_Diagram.png` — generated at `sleep_visualization.R:832`
**Paper use** Methods Fig 1
**What each part means** Vertical flow of 5 stages: Raw Load → Parsed → Algorithmic Correction → Manual Correction → Final Valid. Each box = record count + % of total; two correction stages also show participant counts. **Right-side annotation** breaks the final classification: Not Reported, Clean, Unusual (Accepted), Error (Reviewed), Equal Time, and "% of participants with ≥1 correction". Bottom text = % of raw records entering analysis.
**Healthy look** Flow narrows gently; Clean dominates the annotation; Error + Unusual small (< 5%); correction stages show small counts (only genuinely broken records fixed).
**Anomaly →** Huge Error/Unusual share → AM/PM confusion or bad parsing upstream; near-zero correction counts on data known to contain errors → detection missed rows.
**Precise rules**

- **Data source.** One row per raw diary entry; the five stages are
  Raw Load → Parsed → Algorithmic Correction → Manual Correction → Final Valid.
  Box counts are absolute record counts plus % of `n_total`.
- **The right-side annotation is a `data_category` census of the FINAL output.**
  Every record is assigned exactly one of five classes, in this priority order
  (first match wins, `R/flag_standards.R:40-70`):
  1. `skipped_na` — any of bed/sleep/awake/getup corrected times is missing.
  2. `error` — temporal order broken (`! (bed <= sleep <= awake <= getup)`),
     or zero-length sleep (`sleep == awake`), or adjacent gap > 7 h, or a
     computed sleep span > 24 h. **A zero-length sleep period (sleep == awake)
     is ALWAYS an error**, never Equal Time — a night with zero total sleep is
     a broken record, not a benign pattern.
  3. `equal_time_ok` — order intact AND (`abs(bed − sleep) < 0.01 h` OR
     `abs(awake − getup) < 0.01 h`). The 0.01-hour (~36 s) tolerance absorbs
     floating-point rounding. Three sub-types tracked internally
     (`equal_time_type`): `bed_sleep_equal`, `awake_getup_equal`, `both_equal`.
     These records flow straight through as valid — reporting "went to bed" and
     "fell asleep" at the same clock time is a legitimate diary habit, not a
     quality problem.
  4. `unusual` — order intact, not equal-time, but a gap > 3 h between
     bed→sleep or awake→getup (suspicious but plausible; reviewed, often kept).
  5. `clean` — everything else (order intact, gaps sane).
- **Bottom line** = % of raw records that carry computed sleep metrics
  (`n_valid / n_total`), i.e. everything that was not dropped for missing times.
**FAQ**
FAQ:
- *"Why isn't Equal Time classified as an error?"* Because a zero sleep
  latency is a plausible report. The pipeline reserves `error` for records that
  are internally impossible (order violated, zero-length sleep).
- *"What if BOTH pairs are equal (bed==sleep AND awake==getup)?"* Still
  `equal_time_ok`, tracked as `both_equal`.
- *"What does 0.01 h mean for a reviewer?"* Two timestamps that differ by less
  than ~36 seconds are treated as identical; the tolerance exists only to
  absorb floating-point artifacts, not to hide real differences.
**Paper caption**
*Figure 01. Flow diagram tracing all raw diary
entries through the five pipeline stages (raw load, timestamp parsing,
algorithmic correction, manual correction, final valid output), with record
counts and percentages at each stage. The side annotation reports the final
record classification (clean, unusual accepted, error reviewed, equal-time
benign, not-reported) and the proportion of participants who received at least
one correction. TST = total sleep time; SOL = sleep onset latency.*
#### 02 · Correction Impact

**File** `research_ready/02_Correction_Impact.png` — generated at `sleep_visualization.R:974`
**Paper use** Methods Fig 2
**What each part means** 4 panels: (A) TST delta lollipops — one row per modified record, x = ΔTST min, color orange = algorithmic, blue = manual; (B) same for SOL; (C) identity scatter TST-before vs TST-after, unchanged records at 3% opacity as gray backdrop, diagonal reference line; (D) summary table of TST/SOL mean ± SD before vs after. Caption states how many of how many records were modified.
**Healthy look** Only a small fraction of points deviate from the diagonal; deltas clustered near 0; before/after means nearly identical (non-destructive cleaning).
**Anomaly →** Huge deltas or many rows → corrections are changing real data, not just fixing entry errors; means shift materially → investigate why so much got "fixed".
**Precise rules**

- **Per-record correction status** (`pre_post$status`):
  `manual` if `manually_corrected` is TRUE, else `algorithmic` if `corrected`
  is TRUE, else `none`. A record can never be both — manual wins.
- **Deltas** are defined on the modified records only:
  `ΔTST = tst_after − tst_before` (minutes), where `tst_before` is computed from
  the RAW parsed timestamps as `(awake − sleep) − WASO`, and `tst_after` is the
  pipeline's final `self_diffcalc_totalsleeptime_minutes`. Same structure for
  `ΔSOL = sol_after − sol_before`, with `sol_before = sleep − bed`.
  A negative Δ means the correction shortened the metric.
- **Panel C identity scatter** plots `tst_before` vs `tst_after` for ALL
  records; unchanged records are drawn at 3% opacity (gray) so the ~99% of
  untouched data reads as a faint diagonal backdrop behind the few colored
  corrected points. The dotted 1:1 line is the "no change" reference.
- **Colors**: orange = algorithmic correction, blue = manual correction,
  gray = unchanged.
**FAQ**
FAQ:
- *"Why does most of the data sit on the diagonal?"* Because corrections are
  non-destructive by design: only confirmed input errors are fixed
  (0.58% of records in the real run). Self-report vs. computed discrepancies
  are retained as data, not normalized away.
- *"A point far off the diagonal at 3% opacity — is that a bug?"* No — the
  faint gray includes a few genuinely-modified records too; opacity only
  separates "unchanged mass" visually.
**Paper caption**
*Figure 02. (A) Lollipop plot of
the change in total sleep time (ΔTST, minutes) for each modified record,
colored by correction type (orange = algorithmic, blue = manual). (B) Same for
sleep onset latency (ΔSOL). (C) Identity scatter of TST before vs after
correction; unchanged records are shown at 3% opacity as a gray backdrop.
(D) Summary table of TST and SOL means (± SD) before and after correction.
Corrections are non-destructive: only confirmed input errors were modified.*
#### 02B · Distribution of Sleep Variables

**File** `research_ready/02B_Distribution_Sleep_Variables.png` — generated at `sleep_visualization.R:1018`
**Paper use** Results
**What each part means** Histograms + density curves for key sleep metrics (TST, SOL, WASO, SE) on the **final corrected** data.
**Healthy look** TST 6–8 h peak; SOL right-skewed 10–45; WASO < 60; SE > 85%.
**Anomaly →** SOL flat/bimodal → AM/PM confusion left uncorrected; SE spike at 100% → many all-night-slept records.
**Precise rules**

- **03 (TST distribution)** uses `sleep_duration_h`, defined as the ENHANCED
  TST: sleep period (awake − sleep) minus WASO, i.e.
  `TST = TIB − SOL − WASO`. It is NOT the raw difference between two clock
  times. Mean is drawn as a blue solid line, median as an orange dashed line;
  for skewed sleep data mean > median is expected (long-sleeper tail).
- **02B** plots the same four metrics used by the QC checks — TST, SOL, WASO,
  SE — as histograms with density curves on the final corrected data.
**FAQ**
FAQ:
- *"Why two figures (02B and 03) of distributions?"* 02B is the four-metric
  quality dashboard; 03 isolates TST with mean/median reference lines for the
  Methods section.
**Paper caption**
*Figure 02B. Histograms with density
curves for total sleep time (TST), sleep onset latency (SOL), wake after sleep
onset (WASO), and sleep efficiency (SE) on the final corrected data.*
#### 03 · TST Distribution

**File** `research_ready/03_Sleep_Duration_Distribution.png` — generated at `sleep_visualization.R:1056`
**Paper use** Results
**What each part means** Histogram + density of total sleep time (h) with mean/median reference lines ("Enhanced": sleep period minus WASO).
**Healthy look** Unimodal, 6–8 h center, mean ≈ median.
**Anomaly →** Mean ≫ median → right tail of long sleepers; spike at 0 → zero-duration records.
**Precise rules**

- **03 (TST distribution)** uses `sleep_duration_h`, defined as the ENHANCED
  TST: sleep period (awake − sleep) minus WASO, i.e.
  `TST = TIB − SOL − WASO`. It is NOT the raw difference between two clock
  times. Mean is drawn as a blue solid line, median as an orange dashed line;
  for skewed sleep data mean > median is expected (long-sleeper tail).
- **02B** plots the same four metrics used by the QC checks — TST, SOL, WASO,
  SE — as histograms with density curves on the final corrected data.
**FAQ**
FAQ:
- *"Why two figures (02B and 03) of distributions?"* 02B is the four-metric
  quality dashboard; 03 isolates TST with mean/median reference lines for the
  Methods section.
**Paper caption**
*Figure 03. Histogram and density of TST
(hours) with the mean (blue) and median (orange) marked. TST is computed as the
sleep period minus WASO (TST = TIB − SOL − WASO).*
#### 04 · Sleep Duration vs Time in Bed

**File** `research_ready/04_Sleep_Duration_vs_Time_in_Bed.png` — generated at `sleep_visualization.R:1111`
**Paper use** Results
**What each part means** Scatter TST (h) vs time-in-bed (h) with smooth trend + identity reference.
**Healthy look** Points hug the identity line; spread grows at longer TIB (people lounge in bed); no points with TST > TIB.
**Anomaly →** TST > TIB region populated → duration arithmetic error (WASO subtraction missing); flat cloud → TIB and TST decoupled (parsing problem).
**Precise rules**

- **04** plots TST (h) vs time-in-bed (h). Because TST = TIB − SOL − WASO,
  every valid point must satisfy TST ≤ TIB: points above the identity line are
  physically impossible and indicate a metric-arithmetic problem.
- **04B** plots SOL (h) vs TST (h) with `sol_h` restricted to `[0, 3]` h —
  SOL > 3 h is filtered out purely for visual clarity (those points are
  already flagged as unusual by `max_sol_minutes = 180`). Color = `flag_severity`.
**FAQ**
FAQ:
- *"Why is SOL capped at 3 h here but Figure 13 flags at 120 min?"*
  Same threshold, different unit: 3 h = 180 min = the `max_sol_minutes`
  classification limit. The figure filter just keeps the plot readable.
**Paper caption**
*Figure 04. Scatter of TST (hours) by
time in bed (TIB, hours) with a smoothed trend and identity reference line.
Because TST ≤ TIB by construction, points above the identity line are
physically impossible.*
#### 04B · SOL vs Sleep Duration

**File** `research_ready/04B_SOL_vs_Sleep_Duration.png` — generated at `sleep_visualization.R:1161`
**Paper use** Results
**What each part means** Scatter SOL (h) vs TST (h), **color = data quality**, SOL > 3 h filtered out for clarity.
**Healthy look** Negative correlation (long SOL → short TST); most points SOL < 1 h; quality colors mixed uniformly.
**Anomaly →** Cluster of error/unusual colors at high SOL → sleep-onset mis-entry pattern; vertical stripe at SOL = 0 → zero-latency reporting (equal-time).
**Precise rules**

- **04** plots TST (h) vs time-in-bed (h). Because TST = TIB − SOL − WASO,
  every valid point must satisfy TST ≤ TIB: points above the identity line are
  physically impossible and indicate a metric-arithmetic problem.
- **04B** plots SOL (h) vs TST (h) with `sol_h` restricted to `[0, 3]` h —
  SOL > 3 h is filtered out purely for visual clarity (those points are
  already flagged as unusual by `max_sol_minutes = 180`). Color = `flag_severity`.
**FAQ**
FAQ:
- *"Why is SOL capped at 3 h here but Figure 13 flags at 120 min?"*
  Same threshold, different unit: 3 h = 180 min = the `max_sol_minutes`
  classification limit. The figure filter just keeps the plot readable.
**Paper caption**
*Figure 04B. Scatter of SOL
(hours, restricted to ≤ 3 h for readability) by TST (hours), colored by flag
severity. The negative association (long onset latency, shorter sleep) is the
expected clinical pattern.*
#### 05 · Variability of Sleep Variables

**File** `research_ready/05_Variability_Sleep_Variables.png` — generated at `sleep_visualization.R:1211`
**Paper use** Results
**What each part means** Violin plots + inner boxplots, one per sleep variable, **free Y axis per panel**. Shows distribution shape and spread.
**Healthy look** Roughly symmetric violins for TST; right-skewed SOL/WASO; no panel dominated by spikes.
**Anomaly →** A violin split into two lobes → bimodal behavior (weekday/weekend or 12h-dial); wide flat violin → noisy measurement.
**Precise rules**

- One violin (with inner boxplot) per sleep variable; each panel has a FREE
  Y axis because the variables are on different scales (hours vs %). Comparing
  across panels is about SHAPE (symmetry, spread, bimodality), not absolute values.
**Paper caption**
*Figure 05. Violin plots with overlaid
boxplots for each sleep variable; each panel uses its own y-axis scale, so
panels are compared by distribution shape, not absolute values.*
#### 06 · Sleep Duration Post-Correction

**File** `pipeline_cleaning/06_Sleep_Duration_Post_Correction.png` — generated at `sleep_visualization.R:1285`
**Paper use** Supplement
**What each part means** Density curve of final TST (hours) after manual corrections; one curve over the whole dataset.
**Healthy look** Unimodal, peak 6–8 h, no spikes at 0 or > 12.
**Anomaly →** Flat/bimodal → mixed 12h/24h formats survived parsing; spike at 0 → missing/zero-duration records.
**Precise rules**

- Density of final `sleep_duration_h` after manual corrections are applied.
  Compare against the (auto-detection-only) view in old Figure 14 to see what
  manual review changed. Healthy data: unimodal, 6–8 h mode, no spikes at 0.
**Paper caption**
*Figure 06. Density of final total
sleep time (TST, hours) after all manual corrections were applied. Unimodal
distributions centered on 6–8 h indicate healthy sleep durations; spikes at
zero or > 12 h indicate residual parsing artifacts.*
#### 07 · Flag Composition Stacked

**File** `pipeline_cleaning/07_Flag_Composition_Stacked.png` — generated at `sleep_visualization.R:1350`
**Paper use** Supplement
**What each part means** Stacked histogram: x = sleep duration (h), y = count, **fill = `flag_severity`** (Clean / Minor (1 flag) / Major (2+ flags), not `data_category` — see §6.4). Shows how flag severity varies with sleep duration.
**Healthy look** Clean dominates every duration bin; colored slivers only at extremes (< 3 h, > 12 h) and at 0.
**Anomaly →** Error band wide across mid durations → systematic parsing problem, not extreme-value artifact.
**Precise rules**

- **fill = `flag_severity`**, not `data_category`: `Clean` (0 metric flags),
  `Minor` (1 flag), `Major` (2+ flags). A "flag" is one of
  {SE < 70%, SOL > 1 h, WASO > 1.5 h} (`classification.flag_severity.*`).
  The x-axis is restricted to [0, 16] h for readability; extreme-duration bins
  sit at the edges and are the natural home of Major flags.
**FAQ**
FAQ:
- *"So this is about computed-metric flags, not timestamp errors?"* Correct —
  Figure 13 shows timestamp/error categories; Figure 07 shows how many metric
  flags co-occur at each sleep duration. The two are different classification
  systems (`flag_severity` vs `data_category`).
**Paper caption**
*Figure 07. Stacked histogram of final
sleep duration (hours), with bars colored by flag severity: Clean (no metric
flags), Minor (one flag), Major (two or more flags from {SE < 70%,
SOL > 1 h, WASO > 1.5 h}). Shows whether quality problems concentrate at
extreme durations.*
#### 09 · Bedtime vs Get-up Distribution

**File** `research_ready/09_Bedtime_vs_Getup_Distribution.png` — generated at `sleep_visualization.R:1404`
**Paper use** Results
**What each part means** Two density curves over hour-of-day (0–24): bedtime distribution vs get-up distribution (final corrected times).
**Healthy look** Bedtime peak 22:00–00:00; get-up peak 06:00–08:00; curves don't overlap much.
**Anomaly →** Bedtime peak after 02:00 → delayed-sleep population or PM/AM decode error; get-up peak before 04:00 → mis-parsed evenings.
**Precise rules**

- Two densities over hour-of-day [0, 24): bedtime (`time_bed_corrected`) and
  get-up (`time_getup_corrected`). Uses corrected times, so midnight-flipped
  entries appear at their real evening/morning hours. Peak separation (bed
  ~23:00, get-up ~07:00) is the healthy circadian signature.
**Paper caption**
*Figure 09. Density of bedtime
and get-up clock hours (corrected times) across the day. Peak separation
(evening bedtimes, morning get-ups) is the healthy circadian signature.*
#### 10 · Extreme Sleep Durations

**File** `pipeline_cleaning/10_Extreme_Sleep_Duration.png` — generated at `sleep_visualization.R:1460`
**Paper use** Supplement
**What each part means** Scatter: x = TST (h), y = sleep efficiency (%), restricted to extreme durations (< 4 h or > 10 h). **Color = data quality**, **shape = short vs long sleeper**. Horizontal line at SE = 85% (poor-efficiency threshold).
**Healthy look** Extremes are a small scatter of points; short-sleep points mostly sit at reasonable SE (60–95%); long-sleep points high SE.
**Anomaly →** Cluster of short-sleep points below SE 85% → genuine insomnia vs measurement issue; many points at exactly TST = 0 or 24 → parsing failure.
**Precise rules**

- **Extreme** = TST < 4 h ("short") or TST > 10 h ("long"); everything else is
  "Normal range" and excluded. Point shape encodes short vs long; color encodes
  data quality.
- The horizontal line at **SE = 85%** is the `poor_efficiency_threshold_pct`
  classification boundary (below it a record accumulates a low-SE flag).
**FAQ**
FAQ:
- *"Where do 4 h and 10 h come from?"* They are the plot's own display
  cutoffs, chosen to show the tails; they are looser than the pipeline's
  physiological limits (`duration_extreme`: < 3 h / > 12 h). Don't confuse the
  two sets of thresholds.
**Paper caption**
*Figure 10. Scatter of
total sleep time (TST, hours) versus sleep efficiency (SE, %) restricted to
extreme durations (< 4 h short, > 10 h long). Point shape encodes short vs long
sleep; point color encodes record quality. The horizontal line marks the
SE = 85% poor-efficiency threshold.*
#### 13 · Error Category Distribution

**File** `pipeline_cleaning/13_Error_Category_Distribution.png` — generated at `sleep_visualization.R:1734`
**Paper use** Supplement
**What each part means** Bar chart of auto-detection error/review categories with counts and % labels. Categories mirror `auto_error_desc` prefixes (Temporal / Metrics / Amount / Interval / Timestamp).
**Healthy look** One or two dominant categories (usually Temporal); all categories small in absolute terms.
**Anomaly →** Timestamp or Interval format errors large → participants used non-standard formats en masse; investigate parsing rules.
**Precise rules**

- **Pre-correction auto-detection census**: bars are counts of records flagged
  by the Step 8 check (`checkforerrors_processed`), grouped by `review_source`,
  which is derived from the `auto_error_desc` prefix:
  `[Temporal]` → Temporal Issues, `[Metrics]` → Metrics Issues,
  `[Amount]` → Amount/Input Flags, `[Interval]` → Interval Format Errors,
  `[Timestamp]` → Timestamp Format Errors, else → Other Issues.
- The subtitle records the metric-validation rules that feed it: SOL > 120 min,
  SE < 0 or > 100%, TST/TIB ratio < 0.5.
**FAQ**
FAQ:
- *"Why 'pre-correction'?"* This figure answers "what did the RAW data contain
  that needed fixing?" — the burden the pipeline had to absorb. Figure 18 shows
  the post-correction residual.
**Paper caption**
*Figure 13. Bar chart of records
flagged by the post-correction auto-detection pass, grouped into six review
classes derived from the flag description: temporal issues, metric issues,
amount/input flags, interval format errors, timestamp format errors, and other.
Counts are pre-correction — they show the burden the pipeline had to absorb.*
#### 13B · Adjacent Timestamp Gaps

**File** `pipeline_cleaning/13B_Adjacent_Timestamp_Gaps.png` — generated at `sleep_visualization.R:1817`
**Paper use** Supplement
**What each part means** Histogram of gap hours between adjacent diary events (bed→sleep, sleep→awake, awake→getup), **pre-correction raw times**. Negative gap = the later event's clock time precedes the earlier one. Vertical reference line at 0.
**Healthy look** Mass of gaps in +0.5 to +2 h (positive, sane order); a small negative tail.
**Anomaly →** Large negative mass → widespread order errors the normalizer will flip; bimodal around ±12 h → 12h-dial AM/PM habit.
**Precise rules**

- A **gap** is `later_event − earlier_event` in hours for the three adjacent
  pairs: bed→sleep, sleep→awake, awake→getup, computed on the RAW (decoded)
  timestamps before any correction. **Negative gap** = the later event's clock
  time precedes the earlier event's (e.g. awake 01:00 then sleep 23:00 the same
  row) — the raw signature of an order error or a 12h-dial habit.
- The distribution tells you which failure mode dominates: a negative mass near
  −1 h to −3 h = order swaps (fixed by the swap step); a bimodal ±12 h pattern =
  participants entering PM times as AM (fixed by the 12 h flip, `flip_gap_hours = 12`).
**Paper caption**
*Figure 13B. Histogram of the time
gap (hours) between adjacent diary events (bed→sleep, sleep→awake,
awake→getup) computed from raw, pre-correction timestamps. Negative gaps
indicate the later event's clock time precedes the earlier one, the signature
of order errors and 12 h AM/PM dial habits.*
#### 13C · Detection Outcomes Heatmap

**File** `pipeline_cleaning/13C_Detection_Outcomes_Heatmap.png` — generated at `sleep_visualization.R:1876`
**Paper use** Supplement
**What each part means** Tile heatmap from synthetic error-injection validation: rows/columns = injected error types, cells = detection outcome counts vs modal result. Validates the detector fires on the errors it should.
**Healthy look** Strong diagonal (detected where injected); off-diagonal near zero.
**Anomaly →** Off-diagonal mass → detector misses a class or false-fires on clean controls.
**Precise rules**

- **13C** comes from the SYNTHETIC error-injection benchmark
  (`validation/synthetic/results/detection_outcomes_v4_current.csv`): known
  errors were injected into clean data, and the cell shows the modal detection
  outcome (`CORRECT` / `FLAGGED_UNRESOLVED` / `MISREPAIRED` ...) per injected
  category. The `modal_n` reference is the most common outcome count; rows
  whose count deviates (`n != modal_n`) are the interesting ones. A strong
  diagonal = the detector fires on the error it was injected with.
- **13D** relates cleaning thresholds to measurement noise: the x-axis is
  `threshold / Bland-Altman measurement noise`, and the vertical line marks the
  configured operating point (gap 3 h / flip 12 h). Points right of the line
  are thresholds safely above noise; a detection cliff before the line means
  the threshold sits inside the noise band.
**FAQ**
FAQ:
- *"Why do synthetic-injection figures belong in a QC folder?"* They are the
  evidence that the detection numbers mean something — they validate the
  detector on known ground truth rather than on real data where "correct" is
  unknowable.
**Paper caption**
*Figure 13C. Heatmap from the
synthetic benchmark in which known errors were injected into clean records;
each cell reports the modal detection outcome (correctly auto-fixed,
flagged for review, or misrepaired) for one injected error class. A strong
diagonal indicates the detector fires on the error it was injected with.*
#### 13D · Threshold vs Measurement Noise

**File** `pipeline_cleaning/13D_Threshold_vs_Noise_Ratio.png` — generated at `sleep_visualization.R:1920`
**Paper use** Supplement
**What each part means** Columns showing detection rate against the **threshold-to-Bland-Altman-noise ratio**; vertical line = the chosen operating point. Ties cleaning thresholds to measurement noise rather than arbitrary cutoffs.
**Healthy look** Detection rate high and flat to the right of the vertical line; no cliff at the chosen point.
**Anomaly →** Detection drops before the vertical line → threshold is inside noise band; raise the threshold.
**Precise rules**

- **13C** comes from the SYNTHETIC error-injection benchmark
  (`validation/synthetic/results/detection_outcomes_v4_current.csv`): known
  errors were injected into clean data, and the cell shows the modal detection
  outcome (`CORRECT` / `FLAGGED_UNRESOLVED` / `MISREPAIRED` ...) per injected
  category. The `modal_n` reference is the most common outcome count; rows
  whose count deviates (`n != modal_n`) are the interesting ones. A strong
  diagonal = the detector fires on the error it was injected with.
- **13D** relates cleaning thresholds to measurement noise: the x-axis is
  `threshold / Bland-Altman measurement noise`, and the vertical line marks the
  configured operating point (gap 3 h / flip 12 h). Points right of the line
  are thresholds safely above noise; a detection cliff before the line means
  the threshold sits inside the noise band.
**FAQ**
FAQ:
- *"Why do synthetic-injection figures belong in a QC folder?"* They are the
  evidence that the detection numbers mean something — they validate the
  detector on known ground truth rather than on real data where "correct" is
  unknowable.
**Paper caption**
*Figure 13D. Detection rate
as a function of the threshold-to-measurement-noise ratio, where noise is
quantified from Bland-Altman analysis. The vertical line marks the configured
operating point (3 h gap threshold, 12 h flip threshold); thresholds left of
the line sit within measurement noise.*
#### 17 · Top Participants by Flag Rate

**File** `pipeline_cleaning/17_Top_Participants_Flags.png` — generated at `sleep_visualization.R:2222`
**Paper use** Supplement
**What each part means** Bar chart: top 15 participants by algorithm-flagged record RATE (flags ÷ days), with PID labels.
**Healthy look** Rates modest (< 30%); no single participant dominating.
**Anomaly →** One participant at > 60% → habitual 12h-dial or recurring format issue; check their raw pattern.
**Precise rules**

- **Rate**, not count: `flags / days observed` for that participant. Counts
  reward participants with more diary days; rates identify participants whose
  records are disproportionately flagged. P26 extends this to "worth a second
  look" ranking using `total_days` normalization.
**FAQ**
FAQ:
- *"A high flag rate — is that data error?"* Not necessarily. It flags records
  worth checking by hand; the verdict comes from review, not from the figure.
**Paper caption**
*Figure 17. The 15 participants with the
highest auto-detected flag rate (flags per observed diary day), not raw flag
counts, so participants with few diary days are not overrepresented.*
#### 18 · Auto-Detected Dashboard

**File** `pipeline_cleaning/18_Auto_Detected_Dashboard.png` — generated at `sleep_visualization.R:2322`
**Paper use** Supplement
**What each part means** Left text panel "Key Metrics": total records, flagged records, manually-corrected count. Right: stacked bar of review_source classes (Temporal Issues / Metrics Issues / Amount/Input Flags / Interval Format Errors / Timestamp Format Errors / Other).
**Healthy look** Flagged ≪ total; Temporal largest class; counts match `correction_status_final.csv`.
**Anomaly →** Counts disagree with the CSV → ledger/table pipeline out of sync; Amount flags high → substance-use coding issue.
**Precise rules**

- **Key Metrics panel**: `total records`, `flagged records`, `manually
  corrected count` — these three numbers must agree with
  `output/correction_status_final.csv` (`n_total`, sum of checkforerrors rows,
  `n_corrected`); disagreement means the ledger and the figure pipeline are out
  of sync.
- **Review-source bar**: the same 6 `review_source` classes as Figure 13,
  but on the post-correction flag set.
**Paper caption**
*Figure 18. Key metrics panel (total records,
flagged records, manually corrected records) alongside a stacked bar of the six
review-source classes. Counts must match `correction_status_final.csv`.*
#### 20 · SOL Perception Bias

**File** `research_ready/20_SOL_Perception_Bias.png` — generated at `sleep_visualization.R:2547`
**Paper use** Results
**What each part means** Histogram of |subjective SOL − computed SOL| (minutes) with reference line; shows perception bias magnitude.
**Healthy look** Mass near 0; small tail; mean bias < 30 min.
**Anomaly →** Large systematic shift (mean ≫ 0) → participants misjudge onset latency; bimodal → subset misusing the field.
**Precise rules**

- **20 (SOL):** objective SOL = `time_sleep_corrected − time_bed_corrected`
  (minutes); subjective SOL = the participant's reported
  `duration_totalmin_sol_estimate_am`. `bias = objective − subjective`;
  the histogram shows the absolute difference. A systematic positive shift =
  participants under-report how long it took to fall asleep.
- **20B (WASO):** same construction for wake-after-sleep-onset: reported
  nighttime wakefulness vs computed WASO from the corrected timeline.
**Paper caption**
*Figure 20. Histogram of the absolute difference
between objective SOL (derived from corrected bed and sleep timestamps) and
subjective SOL (participant's reported onset latency), in minutes.*
#### 20B · WASO Perception Bias

**File** `research_ready/20B_WASO_Perception_Bias.png` — generated at `sleep_visualization.R:2602`
**Paper use** Results
**What each part means** Histogram of |self-reported nighttime wakefulness − computed| (minutes); same interpretation as Figure 20 but for WASO.
**Healthy look** Mass near 0, small tail.
**Anomaly →** Large tail → participants under/over-report wake bouts; check WASO duration corrections.
**Precise rules**

- **20 (SOL):** objective SOL = `time_sleep_corrected − time_bed_corrected`
  (minutes); subjective SOL = the participant's reported
  `duration_totalmin_sol_estimate_am`. `bias = objective − subjective`;
  the histogram shows the absolute difference. A systematic positive shift =
  participants under-report how long it took to fall asleep.
- **20B (WASO):** same construction for wake-after-sleep-onset: reported
  nighttime wakefulness vs computed WASO from the corrected timeline.
**Paper caption**
*Figure 20B. Histogram of the absolute difference
between computed wake-after-sleep-onset and the participant's self-reported
nighttime wakefulness, in minutes.*
#### 21 · Substance Use Availability

**File** `research_ready/21_Substance_Use_Availability.png` — generated at `sleep_visualization.R:2674`
**Paper use** Supplement
**What each part means** Bar chart: % of records with data per substance (caffeine, alcohol, nicotine, cannabis).
**Healthy look** High availability (> 80%) for the substances your study asks about; low = expected skip pattern.
**Anomaly →** Near-zero availability for a substance you measure → column mapping or collection failure.
**Precise rules**

- **Units** (from the plotting config): caffeine = **cups**,
  alcohol = **standard drinks**, nicotine = **doses**, cannabis = **doses**.
  Figure 21 shows % of records with data (availability); Figures 22–24 show the
  value distribution (boxplot + jitter for 22; count bars for 23/24).
- Extreme values feed `AMOUNT_FLAG` in Step 8; a spike at an implausible dose
  (e.g. ≥ 10 "cups") usually means a units mix-up in the source export
  (cups vs cans vs servings).
**Paper caption**
*Figure 21. Percentage of records
containing data for each substance (caffeine, alcohol, nicotine, cannabis),
indicating how completely each substance domain was reported.*
#### 22 · Substance Use Value Distribution

**File** `research_ready/22_Substance_Use_Distribution.png` — generated at `sleep_visualization.R:2777`
**Paper use** Supplement
**What each part means** Boxplots + jitter of reported values per substance.
**Healthy look** Compact boxes at plausible doses (caffeine 0–4 cups; alcohol 0–3 drinks); few outliers.
**Anomaly →** Extreme outliers → amount flags (coding units wrong, e.g. drinks vs servings); wide box → non-standard dosing.
**Precise rules**

- **Units** (from the plotting config): caffeine = **cups**,
  alcohol = **standard drinks**, nicotine = **doses**, cannabis = **doses**.
  Figure 21 shows % of records with data (availability); Figures 22–24 show the
  value distribution (boxplot + jitter for 22; count bars for 23/24).
- Extreme values feed `AMOUNT_FLAG` in Step 8; a spike at an implausible dose
  (e.g. ≥ 10 "cups") usually means a units mix-up in the source export
  (cups vs cans vs servings).
**Paper caption**
*Figure 22. Boxplots with jittered points
of reported values per substance. Units: caffeine = cups, alcohol = standard
drinks, nicotine and cannabis = doses.*
#### 23 · Caffeine Consumption

**File** `research_ready/23_Caffeine_Consumption.png` — generated at `sleep_visualization.R:2839`
**Paper use** Supplement
**What each part means** Bar chart of caffeine distribution (cups/day) with counts.
**Healthy look** Right-skewed, mode 0–1 cups.
**Anomaly →** Spike at implausible values (≥ 10) → unit confusion (cans vs cups) feeding `AMOUNT_FLAG`.
**Precise rules**

- **Units** (from the plotting config): caffeine = **cups**,
  alcohol = **standard drinks**, nicotine = **doses**, cannabis = **doses**.
  Figure 21 shows % of records with data (availability); Figures 22–24 show the
  value distribution (boxplot + jitter for 22; count bars for 23/24).
- Extreme values feed `AMOUNT_FLAG` in Step 8; a spike at an implausible dose
  (e.g. ≥ 10 "cups") usually means a units mix-up in the source export
  (cups vs cans vs servings).
**Paper caption**
*Figure 23. Count of records by reported caffeine
consumption (cups/day). Right-skewed with a mode of 0–1 cups is expected;
spikes at implausible values indicate unit confusion.*
#### 24 · Alcohol Consumption

**File** `research_ready/24_Alcohol_Consumption.png` — generated at `sleep_visualization.R:2874`
**Paper use** Supplement
**What each part means** Bar chart of alcohol distribution (drinks/day) with counts.
**Healthy look** Right-skewed, mode 0.
**Anomaly →** Spike at high values → same unit concern as caffeine; verify `alcoholtoday_PM` coding.
**Precise rules**

- **Units** (from the plotting config): caffeine = **cups**,
  alcohol = **standard drinks**, nicotine = **doses**, cannabis = **doses**.
  Figure 21 shows % of records with data (availability); Figures 22–24 show the
  value distribution (boxplot + jitter for 22; count bars for 23/24).
- Extreme values feed `AMOUNT_FLAG` in Step 8; a spike at an implausible dose
  (e.g. ≥ 10 "cups") usually means a units mix-up in the source export
  (cups vs cans vs servings).
**Paper caption**
*Figure 24. Count of records by reported alcohol
consumption (standard drinks/day).*
#### A1 · Step Flag Ledger

**File** `pipeline_cleaning/A1_Step_Flag_Ledger.png` — generated at `figure12_step_flag_table.R:15`
**Paper use** Supplement
**What each part means** Grid table: rows = step × standard × category; columns = count, n_total, n_corrected. Standards: `field_misentry` (1.5), `data_category` (4), `flag_severity` (7), `duration_extreme` (7), `checkforerrors` (8). Read it like the CSV ledger (§2) rendered as a figure.
**Healthy look** Each standard's counts first appear at its step and stay constant after; `equal_time_ok + skipped_na = n_total` from Step 6 on.
**Anomaly →** Counts change after first computation step → pipeline instability (see §2 validation rules).
**Precise rules**

- One row per step × standard × category. The five standards are computed ONCE
  at a fixed step and never recomputed:
  `field_misentry` @ Step 1.5, `data_category` @ Step 4, `flag_severity` @ Step 7,
  `duration_extreme` @ Step 7, `checkforerrors` @ Step 8.
- **Ledger identity (the main sanity check):** from Step 6 onward
  `equal_time_ok + skipped_na = n_total` for `data_category`, and
  `Clean + Minor + Major = n_total − skipped_na` for `flag_severity`.
- See §2 of this vignette for the full validation ruleset.
**Paper caption**
*Figure A1. Per-step × per-standard × per-category record
counts for the five evaluation systems (field mis-entry, data category, flag
severity, duration extreme, check-for-errors). Each standard is computed once
at a fixed pipeline step; counts must remain constant thereafter.*
#### P26 · Participants Worth a Second Look

**File** `pipeline_cleaning/P26_PerParticipant_Flag_Rate.png` — generated at `sleep_visualization.R:2492`
**Paper use** Supplement
**What each part means** Bar chart of participants with the highest flagged-record rate, for manual review prioritization.
**Healthy look** Short list of a few participants; no runaway outlier.
**Anomaly →** Long list → systemic issue affecting many participants, not individual behavior.
**Precise rules**

- **Rate**, not count: `flags / days observed` for that participant. Counts
  reward participants with more diary days; rates identify participants whose
  records are disproportionately flagged. P26 extends this to "worth a second
  look" ranking using `total_days` normalization.
**FAQ**
FAQ:
- *"A high flag rate — is that data error?"* Not necessarily. It flags records
  worth checking by hand; the verdict comes from review, not from the figure.
**Paper caption**
*Figure P26. Ranked bar of participants
by flagged-record rate, for manual review prioritization. A high rate flags
records worth hand-checking; it is not a verdict that the data are wrong.

### 7.2 Publication figures (`research_ready/`)*
#### R25 · Sleep Regularity — Weekday vs Weekend

**File** `research_ready/R25_Sleep_Regularity_Weekday_Weekend.png` — generated at `sleep_visualization.R:2937`
**Paper use** Results
**What each part means** Violin + boxplot of clock hours (bedtime and get-up) split by day type (weekday/weekend).
**Healthy look** Weekend bedtimes ~0.5–1 h later; get-up ~1 h later; distributions otherwise similar.
**Anomaly →** Weekday = weekend exactly → day_type mis-assigned or dates stripped; huge weekend shift → strong social jetlag (report it).
**Precise rules**

- `day_type` is derived from the corrected bedtime: `wday(time_bed_corrected,
  week_start = 1) >= 6 → "Weekend"` (Saturday, Sunday), else "Weekday"
  (Monday–Friday). Violins show the clock hour of bedtime and get-up split by
  day type. A small (≤ 1 h) weekend delay is the healthy signature; a large
  shift is social jetlag worth reporting as a finding.
**FAQ**
FAQ:
- *"Which day does a bedtime at 02:00 belong to?"* The calendar day of the
  `time_bed_corrected` timestamp, which after the midnight-flip correction is
  the evening the participant WENT to bed (e.g. Saturday night 23:30 → wday of
  Saturday... Sunday 02:00 belongs to Saturday's row after normalization).
**Paper caption**
*Figure R25. Violin plots with
boxplots of bedtime and get-up clock hours, split by day type (weekday,
Monday–Friday; weekend, Saturday–Sunday, derived from the corrected bedtime
date). A modest (≤ 1 h) weekend delay is expected; a large shift indicates
social jetlag.*
#### R26 · Sleep Composition — TIB Breakdown

**File** `research_ready/R26_Sleep_Composition_TIB_Breakdown.png` — generated at `sleep_visualization.R:2987`
**Paper use** Results
**What each part means** Stacked bar of time-in-bed composition: TST + SOL + WASO (+ residual) as proportions.
**Healthy look** TST the dominant block (~80–90% of TIB); SOL and WASO small slices.
**Anomaly →** SOL or WASO consuming > 30% of TIB → high sleep fragmentation or onset latency; negative residual → duration arithmetic bug.
**Precise rules**

- **Identity:** `TIB = TST + SOL + WASO` — the figure stacks these three
  components and draws `sol_pct = SOL / total * 100` etc. There is no
  "residual" in the real formula; any apparent gap between the stacked total
  and TIB is rounding. Healthy composition: TST ≈ 80–90% of TIB, SOL and WASO
  small slices.
**Paper caption**
*Figure R26. Stacked bars decomposing time in bed (TIB)
into its components: total sleep time (TST), sleep onset latency (SOL), and
wake after sleep onset (WASO), with TIB = TST + SOL + WASO by construction.
Healthy composition places TST at 80–90% of TIB.*
#### R27 · Sleep Metrics Correlation Matrix

**File** `research_ready/R27_Sleep_Metrics_Correlation_Matrix.png` — generated at `sleep_visualization.R:3016`
**Paper use** Results
**What each part means** Corrplot upper-triangle of pairwise Pearson correlations among TST, SOL, WASO, SE, TIB (final corrected data). Red = negative, green = positive, coefficient printed in each cell.
**Healthy look** Strong expected signs: TST–SE strongly positive, SOL–SE negative, TST–TIB positive, WASO–SE negative.
**Anomaly →** Sign flips (e.g. TST–SE negative) → metric definitions altered or units mixed; near-perfect ±1.0 → two metrics are the same column.
**Precise rules**

- Pairwise **Pearson** correlations (upper triangle) among TST, SOL, WASO, SE,
  TIB on the final corrected data. Red = negative, green = positive, coefficient
  printed per cell.
- **Expected signs:** TST–SE strongly positive (more sleep → higher efficiency,
  by definition SE = TST/TIB); SOL–SE negative; TST–TIB positive (TIB is the
  box TST lives in); WASO–SE negative. A sign flip relative to these is a red
  flag for a metric-definition change or a units mix.
**FAQ**
FAQ:
- *"Why is TST–SE correlation trivially high?"* Because SE is DEFINED as
  TST/TIB — they share a numerator. Treat near-1.0 correlations between
  definitionally-linked metrics as confirmatory, not novel.
**Paper caption**
*Figure R27. Upper-triangle corrplot of
pairwise Pearson correlations among TST, SOL, WASO, SE, and TIB on the final
corrected data (red = negative, green = positive, coefficient per cell).
Expected signs: TST–SE positive, SOL–SE negative, TST–TIB positive,
WASO–SE negative.*

### 6.3 Key definitions

**`equal_time_ok` (Figure 01, `data_category`).** Counts records where two adjacent diary timestamps came out identical but the record is otherwise valid. Fires when bedtime == sleep time (zero sleep latency) **and/or** awake time == getup time (no lingering), **and** the full sequence still satisfies bed ≤ sleep ≤ awake ≤ getup (temporal order intact). Three sub-patterns are tracked internally (`equal_time_type`): `bed_sleep_equal`, `awake_getup_equal`, or `both_equal`. The code treats a ~36-second tolerance (0.01 hours) as "equal" to absorb floating-point rounding (`R/flag_standards.R`).

**The key distinction from the error path:** sleep == awake (a zero-length sleep period) is **never** classified as Equal Time — that is always an error, since a night with zero total sleep duration is a broken record, not a benign reporting pattern. Equal Time is specifically the harmless case where a participant reported two adjacent checkpoints (e.g. "went to bed" and "fell asleep") as the same clock time — a common, legitimate way to fill a sleep diary. Such records flow straight through as valid rather than going to manual review.

**`skipped_na`.** Any record with one or more of the four corrected timestamps missing is classified `skipped_na`, and is excluded from temporal-order evaluation. From Step 6 onward `equal_time_ok + skipped_na = n_total` must hold — this identity is the primary ledger sanity check.

**Unusual vs Error.** `unusual` = an order or gap pattern that is suspicious but plausible (e.g. > 3 h between bed and sleep with sane order) — reviewed, often accepted as-is. `error` = impossible temporal order or zero-length sleep — always reviewed and corrected when confirmed.


## 8. Manual Input Files and How to Run the Pipeline

### 8.1 Manual correction input CSVs

The pipeline reads up to seven human-review CSV files from the working
directory (paths configured under `data.files.*` in the config YAML;
`inst/scripts/00a_setup.R` checks for their presence at startup and reports
missing files). Each file is a REAL participant-data file — keep it out of git
and out of release builds (`.gitignore` + `.Rbuildignore` cover them all).

| CSV (config key) | Consumed at | Why it exists | Created by |
|------------------|-------------|---------------|------------|
| `second_review_checklist.csv` | Step 4.75 (`apply_second_review`) | Locks the second-review consensus for records that were previously reviewed, before corrections are applied | Started by hand; `apply_second_review` appends routed rows |
| `manual_error_corrections.csv` | Step 6 (`step_apply_corrections`) | Human correction decisions for records classified `error` | Step 5 writes `[NEW]manual_error_correction_review.csv`; reviewers fill `new_value_*` and it is saved under this name |
| `manual_unusual_corrections.csv` | Step 6 (`step_apply_corrections`) | Human decisions for records classified `unusual` | Step 5 writes `[NEW]manual_unusual_review.csv`; same review-fill flow |
| `manual_nap_exercise_corrections.csv` | Step 6.5 (`apply_nap_exercise_corrections`) | Corrections to nap and exercise durations | Hand-created; `apply_second_review` may append |
| `manual_sleep_metric_duration_corrections.csv` | Step 6.5 (`apply_sleep_metric_duration_corrections`) | Fixes internal inconsistencies in derived sleep durations | Hand-created |
| `manual_metric_review_acceptances.csv` | Step 6.5 (`apply_metric_review_acceptances`) | Manual accept/reject decisions for computed sleep metrics (TST, SOL, WASO, SE) | Hand-created; `apply_second_review` may append |
| `audit_dispositions.csv` | After Step 9 (`audit_data_integrity`) | Audit ledger recording the disposition of every changed record, for reversibility | Written by the pipeline itself |

**Format:** each file is a plain CSV; the error/unusual files carry
`pid, day_num, row_id, variable, old_value_hhmm, old_value_ampm,
new_value_hhmm, new_value_ampm, correction_type, confidence, reviewer_notes`
(the `[NEW]` review files from Step 5 are pre-filled with diagnostics — fill
the `new_value_*` and `reviewer_notes` columns). Column templates live in
`templates/template_*.csv`; `inst/extdata/stub_*.csv` are shipped in the
package as header-only format examples (no real data).

### 8.2 Run code

**Full pipeline (10 stages + figures):**

```r
library(sleepcleanr)
run_pipeline(config = "real_data_config_fixed.yaml")           # full run
run_pipeline(config = "real_data_config_fixed.yaml",
             skip_visualization = TRUE)                        # cleaning only
```

or, from a shell, `inst/scripts/run.sh` (installs the package if missing,
then runs `run_pipeline()` with the default config).

**Reproducible human-review chain (fasttrack audit):** after a pipeline run,
the candidate-detection → review → rebuild chain is scripted end-to-end with
md5 integrity checks:

```r
source("validation/reproduce_fasttrack_chain.R")   # derive → disambiguate →
                                                  # promote → rebuild; verifies
                                                  # byte-identical outputs
```

**Prerequisites before a run:** the input data file
(`data.files.main`, `.rds` or `.csv`) and the seven manual CSVs of §8.1 must
exist in the working directory; the config YAML points at each. The
`00a_setup` check reports exactly which files are missing rather than failing
silently.
