---
title: "Pipeline Architecture"
description: >
  The 10-step cleaning pipeline, the 8 detection rule families, the decision
  tri-state, and the classification systems — how sleepcleanr is structured.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Pipeline Architecture}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

This vignette documents how the sleepcleanr pipeline is structured: the steps,
the rule families that do the actual cleaning, and the classification systems
that label every record.

```{r setup, eval = FALSE}
library(sleepcleanr)
```

## Pipeline steps

```
Raw Data ──→ Step 1: Load Data ──→ Step 2: Parse Timestamps ──→ Step 3: Parse Intervals ──→ Step 4: Normalize Sequence
                                                                                                      │
                                                                                                      ▼
                                                                                             Step 5: Classify Records
                                                                                             (generates review CSVs)
                                                                                                      │
                                                                                             Step 5.75: Second Review
                                                                                                      │
                                                                                             Step 6: Apply Manual Corrections
                                                                                             (reads manual_error_corrections.csv)
                                                                                                      │
                                                                                             Step 6.5: Apply Duration Corrections
                                                                                             (nap, exercise, SOL/WASO corrections)
                                                                                                      │
                                                                                             Step 7: Compute Sleep Metrics
                                                                                             (TST, SOL, WASO, SE, TIB)
                                                                                                      │
                                                                                             Step 8: Auto-Detect Remaining
                                                                                             (TIMESTAMP/DURATION/AMOUNT/SELF-REPORTED)
                                                                                                      │
                                                                                             Step 8.5: Cross-Participant Check
                                                                                                      │
                                                                                              Step 9: Generate Figures
```

**10 steps** (source: `inst/steps.yaml`):

| Step | Label | Description |
|------|-------|-------------|
| 1 | Load data | .rds/.csv auto-detected; schema validated; optional supplementary file merged |
| 1.5 | Field-misentry check | SOL/WASO clock-time vs duration-field misentry detection on raw data |
| 2-4 | Parse & normalize (S3 chain) | Parse timestamps → parse intervals → normalize sequence |
| 5 | Classify records | Generate manual review CSVs for human approval |
| 5.75 | Second-review consensus | Apply second-review checklist consensus |
| 6-7 | Correct & compute metrics (S3 chain) | Manual + duration corrections; TST/SOL/WASO/SE metrics; has_correction enum |
| 8 | Auto-detect remaining issues | TIMESTAMP/DURATION/AMOUNT/SELF-REPORTED flag classification |
| 8.5 | Cross-participant consistency check | Global consistency audit across participants |
| 9 | Generate diagnostic figures | 30 figures (14 QC + 16 research) + figure_index.png contact sheet + RUN_INFO.txt |
| 10 | Build delivered datasets | finalize_columns() selects/renames to Dataset A/B per column dictionary |

## Detection rule families

The cleaning logic is organized into eight rule families. Each family targets
one failure mode of free-text self-report, has a defined decision behavior
(see tri-state below), and is validated against a synthetic error class that
exercises it (see the validation-methodology vignette). This is the part of
the pipeline that does the actual work — the steps above are the scaffolding
around it.

| Family | Failure mode targeted | Decision on hit |
|---|---|---|
| **Timestamp standardization** | Non-uniform raw strings (`10.30`, `7:30`, bare hours, hms/difftime types) that can silently misparse (e.g. minutes read as seconds) | FLAG (format risk only) |
| **Temporal order validation** | Bedtime entered after sleep onset; AM/PM flips near midnight; cross-day confusion | AUTO_FIX (needs corroboration) / FLAG |
| **Free-text duration parsing** | SOL/WASO typed as text (`90:00` meaning 90 min, `0130`, `p` for 0) | FLAG (clock-form never auto-corrected) |
| **Internal consistency check** | Derived SOL contradicts self-reported SOL; derived duration exceeds the bed→sleep window | AUTO_FIX component / AUDIT |
| **Redundancy-confirmation check** | A "correction" that moves values *away* from self-report (silent worsening) | **Veto authority over AUTO_FIX** |
| **Cross-day stability screening** | Implausible day-to-day SOL/WASO jumps within a participant | AUDIT-only (may be real signal) |
| **Correction provenance audit** | Human correction notes no code path understands (blind spots) | AUDIT-only |
| **Post-correction verification** | Corrected timestamps still disagree with reported durations | FLAG / PASS (release gate) |

**Decision tri-state:**

| Decision | Meaning |
|---|---|
| **AUTO_FIX** | Deterministic, reversible correction applied without human review. For temporal order, requires corroboration: order violation ∧ internal consistency ∧ redundancy confirmation. |
| **FLAG** | Sent to the human review queue (CSV workflow); no automatic action. |
| **AUDIT-only** | Counted and reported; never modifies data. |

Every rule family is deterministic (same input → same output, snapshot-verified),
which is why inter-rater reliability statistics do not apply — there is no
rater variance. Human co-review agreement is reported instead (see the
validation-methodology vignette, Step 7).

## Classification systems

| System | Source | Categories |
|--------|--------|------------|
| `data_category` | Step 6 (temporal) | clean, error, unusual, equal_time_ok, skipped_na |
| `has_correction` | Step 7 (traceability) | none, algorithmic, manual, both |
| `flag_severity` | Step 7 (metrics) | Clean, Minor (1 flag), Major (2+ flags) |
| `checkforerrors_summary` | Step 8 (auto-detect) | TIMESTAMP_ISSUE, DURATION_ISSUE, AMOUNT_FLAG, SELF_REPORTED_FLAG, CLEAN |

## Figures

Generated figures are split into two folders so pipeline-quality views
(every run) and analysis views (only when the study contains the relevant
variables) do not get confused.

| Folder | Content |
|--------|---------|
| `pipeline_cleaning/` | Pipeline flow diagram, data quality dashboard, flag composition, per-participant flag rate, step flag ledger |
| `research_ready/` | Correction impact (before/after delta), sleep variable distributions, perception bias, substance use, sleep regularity, correlation matrix |

## Non-destructive model

**The pipeline never deletes a record.** Cleaning means *adding labels and
corrected columns*, not removing rows. The number of records in
`corrected_ema_data` always equals the number of records you put in.

```
Input:  N raw records
Output: N records, each with additional classification columns
        (data_category, flag_severity) and corrected-value columns
        (time_*_corrected). Raw columns are preserved untouched.
```

- **Nothing is lost.** Raw timestamps (`time_bed_am_hhmm_ampm`, etc.) stay in
  the data. Corrected versions are added as new columns
  (`time_bed_corrected`), so every correction is auditable against the
  original value.
- **Every record is labelled, not removed.** Records that fail validation are
  tagged (`data_category = "error"` or `"unusual"`) but remain in the dataset.
  You decide later whether to include or exclude them in analysis.
- **"Final Clean Dataset" is a recommended analysis subset, not a physically
  separate file.** It is the records where `data_category` is `clean` or
  `equal_time_ok`:

```r
clean_data <- corrected_ema_data[corrected_ema_data$data_category %in% c("clean", "equal_time_ok"), ]
```

**One exception:** records with all four sleep timestamps missing
(`data_category = "skipped_na"`) stay in the file, but their TST/SOL/WASO
metrics are `NA` — sleep metrics cannot be computed without timestamps. They
are available for compliance/attrition analyses (e.g. "how many days did each
participant complete") but not for sleep-metric analyses.

| Record type | Still in the data? | Has sleep metrics? | Suitable for sleep analysis? |
|------------|:---:|:---:|:---:|
| clean / equal_time_ok | Yes | Yes | **Yes** (recommended) |
| error / unusual (flagged) | Yes | Yes | With caution — researcher's decision |
| skipped_na (no timestamps) | Yes | No (NA) | No — compliance analysis only |
