---
title: "Testing Coverage"
description: >
  The 190+ testthat expectations across 12 test files: what the tests
  verify, how to run them, and how snapshot verification works.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Testing Coverage}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

The pipeline includes 190+ testthat expectations across 12 test files, all
exercising software correctness — does the code do what it was designed to
do — as distinct from methodological validity, which is covered in the
validation-methodology vignette.

```{r setup, eval = FALSE}
library(splsleep)
```

## Test files and coverage

| Test File | Coverage |
|-----------|----------|
| `test-normalize.R` | AM/PM correction, minor order errors, midnight crossing, edge cases |
| `test-interval.R` | Colon-format edge cases in duration parsing |
| `test-pipeline.R` | End-to-end run on synthetic data, config loading, column adaptation |
| `test-sleep-diary.R` | S3 construction, validation, coercion, generics, step-contract assertion, provenance |
| `test-flag-standards.R` | Flag evaluators for field misentry, data category, duration extreme, flag severity, checkforerrors; step-ledger logging |
| `test-correction-engine.R` | Every classification outcome (order/bed-sleep/awake-getup/24h errors, equal-time, unusual, skipped-NA, multiple errors, suspicious-latency flags) |
| `test-classification-thresholds.R` | Boundary behavior at every classification threshold (7h error, 3h/15h unusual, suspicious-latency) |
| `test-auto-detection-thresholds.R` | SOL/SE/TST-TIB boundary behavior for the Step 8 auto-detection flags |
| `test-finalize-columns.R` | Dictionary ↔ delivered-column consistency, A/B join key uniqueness, unit transforms, reserved-column pass-through, export guard, missing/optional-column handling |
| `test-nonfinite-guards.R` | NA/Inf handling in duration and flag-severity evaluators (regression tests for two real bugs) |
| `test-config-data.R` | Config file loading (RDS/CSV, legacy keys, column mapping, friendly error messages) |
| `test-script-copies-in-sync.R` | Root and `inst/scripts/` copies of every dual-maintained script stay byte-identical |

## How to run the tests

```r
# Installed-package mode (fast, tests the built package):
testthat::test_package("splsleep")

# Source-development mode (tests the working tree; use this when editing
# code — test_dir() alone does NOT load the package, so load it first):
pkgload::load_all(".")
testthat::test_dir("tests/testthat")

# Preferred dev workflow (loads source + runs tests in one call, needs
# devtools installed):
devtools::test()
```

Full package check (documentation, examples, tests, and the
`verify_reference_fidelity` / dual-copy checks):

```r
devtools::check()   # needs devtools; equivalently: R CMD check splsleep_*.tar.gz
```

`devtools` is a development-only tool — it is intentionally **not** a
dependency of the package (not in `DESCRIPTION`), so it is never required to
install or run splsleep itself.

## Snapshot verification

Snapshot verification (`inst/verification/`, `verify_v1_3_snapshot.R`)
confirms the current S3 pipeline chain produces byte-identical output to the
legacy pipeline path on real data. `verify_reference_fidelity.R` separately
pins each of the 8 core metric formulas against a documented baseline
(`--strict` mode is CI-wired).

## 中文 — 测试覆盖率

管线包含 12 个测试文件、190+ 条 testthat 断言，全部检验软件正确性——代码是否
按设计跑——区别于方法学效度（后者见验证方法学 vignette）。测试覆盖：AM/PM 校正
与边界、冒号格式、端到端运行、S3 构造与步骤契约、5 个 flag 评估器、全部分类
结果、阈值边界、自动检测边界、finalize 交付契约、NA/Inf 守卫（两个真实 bug 的
回归）、配置加载、双维护脚本字节一致性。

**运行方式**：`testthat::test_package("splsleep")`（已装包，快）/
`pkgload::load_all(".")` + `testthat::test_dir("tests/testthat")`（源码开发）/
`devtools::test()`（推荐）。完整包检查：`devtools::check()`。`devtools` 是纯
开发工具，刻意不是包依赖。

**Snapshot 验证**（`verify_v1_3_snapshot.R`）：当前 S3 管线链与旧管线路径在
真实数据上输出字节一致（13/13）。`verify_reference_fidelity.R` 单独把 8 个核心
指标公式钉在文档化基线上（`--strict` 模式已接入 CI）。
