---
title: "R01_sleepdiaryema_cleaning"
author: "Maia ten Brink"
date: "5/10/2022"
output: html_document
---

Last updated: Maia, 02/07/2025

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
rm(list=ls())
```

## Outline

### Completed

Link IDs and date of observation to release EMA dataset
Run function process_time to do first-pass cleaning on time variables of interest
processes both timestamps and intervals
identifies and fixes AM/PM errors
Saves time processed data
Creates separate dataframes for AM, PM, and midday pings


### Remains to be completed for data share 2025

### Remains to be completed for later data share: processing and cleaning timestamps - Maia do this task

Preprocess the time variables - check what “still under construction” means??
Calculate calculated (not self-reported) sleep vars (see calculate_sleep_time_vars.R) and creates checks in find_time_errors.R (not finished) (e.g. distribution plots, outliers, etc) to identify potential time entry errors: the goal is to create this in such a way that future me and other researchers can use it as a general tool:
negative time duration
far outside typical distribution
AM/PM entry errors (already mostly corrected in first pass cleaning)


## Load filepaths

This sources custom code and filepaths.
```{r}
source(here::here("file_paths.R"))

source(here::here("R", "calculate_sleep_time_vars.R"))
source(here::here("R", "separate_ping_occasions.R"))

source(here::here("R", "process_timestamp_emadatarelease.R"))
source(here::here("R", "process_interval_emadatarelease.R"))

#Archived:
# source(here::here("R", "process_time.R"))
# source(here::here("R", "process_time_emadatarelease.R"))
# source(here::here("R", "process_timestamp.R"))
# source(here::here("R", "process_interval.R"))
```

## Load libraries

If you are just running locally, you can also load packages with this code.
```{r libraries}
# install.packages("devtools")
# devtools::install_github("amehtaSF/rsurveyutils")

library(rsurveyutils)
library(tidyverse)
library(tidylog)
library(here)
library(broom)
library(janitor)
library(stringi)
library(stringr)
library(furrr)
```

## Load files

```{r Load final_EMA_data_release}
ema_data_release <- read_csv(file = paste0(dir_data, "/processed/from_data_release/sber_ema_anon_20240429.csv"))
```

# Recode the time variables

Note: This is still under construction!

**Timestamp variables:**
time_bed_am
time_sleep_am
time_awake_am
time_getup_am
caffeinetoday_PM
alcoholtoday_PM
nicotine_amount_pm
cannabis_amount_pm

**Time interval variables:**
duration_totalmin_sol_estimate_am (hh:min --> total minutes)
duration_totalmin_waso_estimate_am (hh:min --> minutes)
duration_totalmin_napstoday_PM (hh:min --> minutes)
exercisetoday_PM_totalmin_Light (hh:min --> minutes)
exercisetoday_PM_totalmin_Moderate (hh:min --> minutes)
exercisetoday_PM_totalmin_Vigorous (hh:min --> minutes)
exercisetoday_PM_totalmin_Strength (hh:min --> minutes)

**Unknown format timestamp variables:**
medicines_x_time
acti_time_remove
vital_time_remove
= For now, these variables will remain unprocessed until someone wants to make use of the data

## Run function `process_time` to do first-pass cleaning on time variables of interest:

```{r}
tstamp.vars.to.proc = c("time_bed_am",
"time_sleep_am",
"time_awake_am",
"time_getup_am",
"caffeinetoday_PM",
"alcoholtoday_PM",
"nicotine_amount_pm",
"cannabis_amount_pm")

interval.vars.to.proc = c("duration_totalmin_sol_estimate_am", 
"duration_totalmin_waso_estimate_am",
"duration_totalmin_napstoday_PM",
"exercisetoday_PM_totalmin_Light",
"exercisetoday_PM_totalmin_Moderate",
"exercisetoday_PM_totalmin_Vigorous",
"exercisetoday_PM_totalmin_Strength")
```

Do time processing on released EMA data:

```{r}
#Create copies of df to save all processed time variables
ema_data_release_timeproc <- ema_data_release %>% 
	#rename so that time processing scripts work
	rename(duration_totalmin_sol_estimate_am = duration_sol_estimate_am_hhmm,
				 exercisetoday_PM_totalmin_Light = exercisetoday_PM_Light,
				 exercisetoday_PM_totalmin_Moderate = exercisetoday_PM_Moderate, 
				 exercisetoday_PM_totalmin_Vigorous = exercisetoday_PM_Vigorous,
				 exercisetoday_PM_totalmin_Strength = exercisetoday_PM_Strength) %>% 
	#remove the trailing :00 from interval duration 
	mutate(across(
    all_of(interval.vars.to.proc),
    ~ gsub(":00$", "", .x)
  )) %>%
	#change l,C to PM, AM so that time processing scripts work
	mutate(across(ends_with("_ampm"), 
                ~ case_when(
                  . == "l" ~ "PM",
                  . == "C" ~ "AM",
                  TRUE ~ .  # Keeps NA, AM, and PM unchanged
                )))

#check 
names(ema_data_release_timeproc)
```

Process timestamp and interval columns:
```{r}
#Process timestamp variables
for (varname.count in 1:length(tstamp.vars.to.proc)) {
	ema_data_release_timeproc <- process_timestamp(df = ema_data_release_timeproc, varname = tstamp.vars.to.proc[varname.count], format = "timestamp")
}
rm(varname.count)

#Process interval variables -- Maia's old, dysfunctional code
# for (varname.count in 1:length(interval.vars.to.proc)) {
# 	ema_data_release_timeproc <- process_interval(df = ema_data_release_timeproc, varname = interval.vars.to.proc[varname.count], format = "interval_hhmm")
# }
# rm(varname.count)

#Process interval variables -- Cyra's new, functional code = NOTE: requires /cyra_code/process_interval
for (varname.count in 1:length(interval.vars.to.proc)) {
  cat("Processing:", interval.vars.to.proc[varname.count], "\n")
  ema_data_release_timeproc <- process_interval(df = ema_data_release_timeproc, varname = interval.vars.to.proc[varname.count], format = "interval_hhmm")
}
rm(varname.count)
```

Plan: Cyra will identify where in the process_interval and process_timestamp function these are going wrong, using deidentified data
```{r}
#de-identified for Cyra 10/02/25 (intervals); 11/13/25 (all)
ema_data_release_timeproc %>%
	select(pid, day_num, survey_type, all_of(interval.vars.to.proc), all_of(starts_with(tstamp.vars.to.proc))) %>%
	saveRDS(., file = paste0(dir_data, "/processed/from_data_release/deidentified_intervalvars_forCD_111325.rds"))

cyra_df <- readRDS(file = paste0(dir_data, "/processed/from_data_release/deidentified_intervalvars_forCD_111325.rds"))
```

```{r}
ema_data_release_timeproc %>%
	dplyr::select(pid, ends_with("mincalc"), ends_with("correctionsmade"), ends_with("errors"), ends_with("_hhmm_ampm")) 
```


Save time processed files:

```{r}
save(ema_data_release_timeproc, file = paste0(dir_data, "/processed/time_processed_ema_data_release.RData"))

load(file = paste0(dir_data, "/processed/time_processed_ema_data_release.RData"))
```

## Create calculated variables to compare with self-report time estimates:

Relevant variables:
	# time_bed_am_hhmm_ampm
	# time_sleep_am_hhmm_ampm
	# time_awake_am_hhmm_ampm
	# time_getup_am_hhmm_ampm
	# num_waso_estimate_am
	# duration_totalmin_sol_estimate_am_mincalc
	# duration_totalmin_waso_estimate_am_mincalc
	# num_waso_estimate_am

```{r}
#check that we have the corrected variables

ema_data_release_timeproc %>%
	select(ends_with("hhmm_ampm"))

ema_data_release_timeproc %>%
	select(duration_totalmin_sol_estimate_am, duration_totalmin_waso_estimate_am, duration_totalmin_napstoday_PM)
#conversion of duration in min is missing hours to min
#timestamp correction hasn't happened all the way yet for the naps duration

ema_data_release_timeproc %>%
	select(ends_with("_mincalc"))

ema_data_release_timeproc %>%
	select(ends_with("_checkforerrors"))
```

Use the function `calculate_sleep_time_vars` (currently under construction).

```{r}
ema_data_release_timecalc <- calculate_sleep_time_vars(AM_rawdata = ema_data_release_timeproc)
```

```{r}
#check that it ran
ema_data_release_timecalc %>%
	select(pid, StartDate, starts_with("time"), starts_with("self_diffcalc"))
```


```{r}
save(ema_data_release_timecalc, file = paste0(dir_data, "/processed/time_processed_calculated_emadatarelease.RData"))
```

```{r}
ema_data_release_timecalc <- load(paste0(dir_data, "/processed/time_processed_calculated_emadatarelease.RData"))
```


# Take a look at entry errors rows and make manual corrections:

Calculate mean and SD of all vars:

Timestamps:
	# time_bed_am_hhmm_ampm
	# time_sleep_am_hhmm_ampm
	# time_awake_am_hhmm_ampm
	# time_getup_am_hhmm_ampm

Duration:
	# duration_totalmin_sol_estimate_am_mincalc
	# duration_totalmin_waso_estimate_am_mincalc

Number:
	# num_waso_estimate_am

```{r}
names(ema_data_release_timecalc)

check_for_errors <- ema_data_release_timecalc %>% 
	dplyr::select(pid, date_of_obs, survey_type, survey_comments, survey_concerns, contains("checkforerrors")) %>% 
	filter(if_any(ends_with("checkforerrors"), ~ is.na(.) == FALSE)) %>% #will still come out a true
	## alternative AI suggestions: filter(if_any(ends_with("checkforerrors"), ~ . == TRUE))
	pivot_longer(cols = any(ends_with("checkforerrors")), names_to = "variable_check", values_to = "true_false") %>% #not working!
	## alternative AI suggestions: pivot_longer(cols = ends_with("checkforerrors"), names_to = "variable_check", values_to = "true_false")
	filter(true_false == TRUE)
```

!!!start here: this isn;t working
```{r}
summary_timestamps <- ema_data_release_timecalc %>%
# 	mutate(across(c(time_bed_am_hhmm_ampm, time_sleep_am_hhmm_ampm, 
#                   time_awake_am_hhmm_ampm, time_getup_am_hhmm_ampm), 
#                 ~ hms::as_hms(strptime(.x, format = "%I:%M %p")))) %>% 
	summarise(across(c(time_bed_am_hhmm_ampm, time_sleep_am_hhmm_ampm, 
                  time_awake_am_hhmm_ampm, time_getup_am_hhmm_ampm), 
                ~ mean(na.rm = T))) %>% 
	summarise(across(c(time_bed_am_hhmm_ampm, time_sleep_am_hhmm_ampm, 
                  time_awake_am_hhmm_ampm, time_getup_am_hhmm_ampm), 
                ~ sd(na.rm = T)))	

#summarise(across(.cols = -ID, ~ mean))

ema_data_release_timecalc %>% select(pid, survey_type, time_bed_am_hhmm_ampm)
```


### Plot TIB, TST, SE




!!! there are several errors with time in bed / time asleep 3AM switching to 15:00, or 1AM to 13:00!
fix!!!

```{r}
data_recoded_AM_diaries %>%
	select() %>%
ggplot(., aes(x = diaryday, y = self_diffcalc_sol_minutes, color = id)) +
	geom_line() +
	theme_minimal() +
	theme(legend.position = "none")


```

	#Suffixes
	##Corrected timestamps
	timestamp_suffix <- "_hhmm_ampm"
	##Corrected interval durations in minutes
	interval_suffix <- "_mincalc"
	##Check for errors
	errorcheck_suffix <- "_checkforerrors"

```{r}

```



## Comments and corrections:

```{r}
unique(data_raw$comments_am)
unique(data_raw$survey_comments)
```

Plan: Ashish and Maia split

Go through comments, mark if
sick
traveling
unusual circumstance sleep-related
unusual circumstance other

Corrections to survey responses:
Mark that response has been manually corrected
Make manual correction


# Check situations

Plan:
Spot check % of situations early in EMA, 1x/day
